Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-04-21 manu-draft.qmd
first cut of abstract

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "EBASE manuscript"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism.  Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen concentrations, which can lead to poorly constrained errors that influence the interpretation of results. In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems. This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, or the Estuarine BAyesian Single-station Estimation (EBASE) method.  EBASE provides several advantages over existing models by 1) using a linear relationship between irradiance and production, 2) removing the temperature-dependency of respiration, and 3) applying a wind-based gas exchange formulation method applicable to estuaries.  Moreover, the approach uses a Bayesian framework where critical model parameters are fit as a function of the data and informed using a set of prior distributions.  We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results. The approach is also bench-marked with more conventional open-water methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration). The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.
